# Advanced Plotting 

"The greatest value of a picture is when it forces us to notice what we never expected to see"
-John Tukey

---

## Setup

**1. Install the `tidyverse` package.**

```{r}
library(tidyverse)
```

**2. Filter data.**

We will be using the publication dataset that we read into R in Chapter 2 as `counts`. 

```{r, include = FALSE}
counts <- read.delim("data/counts-raw.txt")
```

```{r}
research <- filter(counts, articleType == "Research Article")
```

## Review of ggplot2 basics

`ggplot2` is a plotting package that makes it simple to create complex plots from data in a data frame. Graphics are built step by step by adding new elements. Adding layers in this fashion allows for extensive flexibility and customization of plots.

A plot can be divided into different fundamental parts:

**Plot = data + aesthetics + geom**

Required building blocks:

* data
* aesthetics - describe how data are mapped to colour, size, shape, location
* geoms - geometric objects like points, lines, shapes

Optional building blocks:

* facets - describes how panel plots should be constructed
* stats - statistical transformations like binning, quantiles, smoothing
* coordinates - describes the system in which the locations of the geoms will be drawn
* scales - what scale an aesthetic map uses (ex. male = red, female = blue)

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```{r, eval = FALSE}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()
```
1. Specify which data set to use for the plot using the `data` argument.

2. Define a "mapping" (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

3. Add "geoms" – graphical representations of the data in the plot (points, lines, bars). `ggplot2` offers many different geoms; common ones include:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for boxplots.
* `geom_histogram()` for histograms.
* `geom_barplot()` for barplots.
* `geom_line()` for trend lines, time series, etc. 

Example:

```{r}
p <- ggplot(research, aes( x = pdfDownloadsCount, y = wosCountThru2011)) + geom_point()
p
```

Adding aesthetics:

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount,
                          y = wosCountThru2011)) +
  geom_point(aes(color = journal))
p
```

## Statistics

The function `geom_smooth()` adds a loess curve to the data along with a 95% confidence interval.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount,
                          y = wosCountThru2011)) +
  geom_point(aes(color = journal)) +
  geom_smooth()
p
```

If we move the colour to the base ggplot call, we get loess curves for each level of that factor.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, 
                          y = wosCountThru2011, 
                          color = journal)) + 
  geom_point() + 
  geom_smooth()
p
```

Check the help page for the function `geom_smooth()` for more information about how the curve is made. For example, to map a linear model onto the plot, you can choose `method = "lm"`.

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount, 
                          y = wosCountThru2011, 
                          color = journal)) + 
  geom_point() + 
  geom_smooth(method = "lm")
p
```

## Scales

Now let's look at the relationship between days since published and citation count.

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = wosCountThru2011)) + 
  geom_point(aes(color= journal))
p
```

It looks like most of the citation counts are close to 0. We can quickly check the distribution of this variable using a qplot.

```{r}
qplot(data = research, x = wosCountThru2011)
```

To control the plot axes, we use variants of the functions `scale_x_*` and `scale_y_*`.

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = wosCountThru2011)) + 
  geom_point(aes(color= journal)) +
  scale_y_log10()
p
```

How can we fix this?

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal))
p
```

Notice what this fix has done to the way the y-axis is labelled. To manually update the axis label, we can use the `scale_y_continuous()` function.

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal)) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100, 1000))
p
```

## Faceting

There are two functions to control how plots are divided into facets: `facet_wrap()` and `facet_grid()`. 



## Themes

In addition to `theme_bw()`, which changes the plot background to white, `ggplot2` comes with several other themes which can be useful to quickly change the look of your visualization. The complete list of themes is available at https://ggplot2.tidyverse.org/reference/ggtheme.html. `theme_minimal()` and `theme_light()` are popular, and `theme_void()` can be useful as a starting point to create a new hand-crafted theme.

The `ggthemes` package provides a wide variety of options (including an Excel 2003 theme). The `ggplot2` [extensions website](https://www.ggplot2-exts.org) provides a list of packages that extend the capabilities of `ggplot2`, including additional themes.

## Color palettes

You can create your own color palettes using the `colorRamp()` or `colorRampPalette()` function. 

`colorRamp()` returns a function that takes values between 0 and 1, indicating the extremes of the color palette.

`colorRampPalette()` returns a function that takes integer arguments and returns a vector of colours. 

```{r}
cols <- colorRamp(c("red", "blue"))
cols(0)
cols(0.5)
cols(1)

cols <- colorRampPalette(c("red", "blue"))
cols(2)
cols(10)
```

Or, you can use the `RColorBrewer` package to get a premade colour palette.

There are three types of palettes:
* Sequential
* Diverging
* Qualitative

```{r, fig.height = 8}
library(RColorBrewer)
display.brewer.all()
```

## Multiple plots

There are two useful packages for combining multiple plots: `gridExtra` and `cowplot`. 

`gridExtra` has two useful functions: `grid.arrange()` and `arrangeGrob()`. However, these functions make no attempt at aligning the plot panels; instead, the plots are simply placed into the grid as they are, so the axes are not aligned. If axis alignment is required, the `plot_grid()` function of the `cowplot` package is better. We will try using both here. 

### The `gridExtra` package

### The `cowplot` package

### Saving plots

The easiest way to save a plot is using the `ggsave()` function. 


## Additional Resources

http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization

http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html


* Mailing list: http://groups.google.com/group/ggplot2
* Wiki: https://github.com/hadley/ggplot2/wiki
* Website: http://had.co.nz/ggplot2/
* StackOverflow: http://stackoverflow.com/questions/tagged/ggplot

Cheatsheet



