# Advanced Plotting 

"The greatest value of a picture is when it forces us to notice what we never expected to see"
-John Tukey

---

## Setup

**1. Install the `tidyverse` package.**

```{r}
library(tidyverse)
```

**2. Filter data.**

We will be using the publication dataset that we read into R in Chapter 2 as `counts`. 

```{r, include = FALSE}
counts <- read.delim("data/counts-raw.txt")
```

```{r}
research <- filter(counts, articleType == "Research Article")
```

## Review of ggplot2 basics

`ggplot2` is a plotting package that makes it simple to create complex plots from data in a data frame. Graphics are built step by step by adding new elements. Adding layers in this fashion allows for extensive flexibility and customization of plots.

A plot can be divided into different fundamental parts:

**Plot = data + aesthetics + geom**

Required building blocks:

* data
* aesthetics - describe how data are mapped to colour, size, shape, location
* geoms - geometric objects like points, lines, shapes

Optional building blocks:

* facets - describes how panel plots should be constructed
* stats - statistical transformations like binning, quantiles, smoothing
* coordinates - describes the system in which the locations of the geoms will be drawn
* scales - what scale an aesthetic map uses (ex. male = red, female = blue)

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```{r, eval = FALSE}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()
```
1. Specify which data set to use for the plot using the `data` argument.

2. Define a "mapping" (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

3. Add "geoms" – graphical representations of the data in the plot (points, lines, bars). `ggplot2` offers many different geoms; common ones include:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for boxplots.
* `geom_histogram()` for histograms.
* `geom_barplot()` for barplots.
* `geom_line()` for trend lines, time series, etc. 

Example:

```{r}
p <- ggplot(research, aes( x = pdfDownloadsCount, y = wosCountThru2011)) + geom_point()
p
```

Adding aesthetics:

```{r}
p <- ggplot(research, aes(x = pdfDownloadsCount,
                          y = wosCountThru2011)) +
  geom_point(aes(color = journal))
p
```

## Statistics

The function `geom_smooth()` adds a loess curve to the data along with a 95% confidence interval.

```{r, message = FALSE}
p <- ggplot(research, aes(x = pdfDownloadsCount,
                          y = wosCountThru2011)) +
  geom_point(aes(color = journal)) +
  geom_smooth()
p
```

If we move the colour to the base ggplot call, we get loess curves for each level of that factor.

```{r, message = FALSE}
p <- ggplot(research, aes(x = pdfDownloadsCount, 
                          y = wosCountThru2011, 
                          color = journal)) + 
  geom_point() + 
  geom_smooth()
p
```

Check the help page for the function `geom_smooth()` for more information about how the curve is made. For example, to map a linear model onto the plot, you can choose `method = "lm"`.

```{r, message = FALSE}
p <- ggplot(research, aes(x = pdfDownloadsCount, 
                          y = wosCountThru2011, 
                          color = journal)) + 
  geom_point() + 
  geom_smooth(method = "lm")
p
```

## Scales

Now let's look at the relationship between days since published and citation count.

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = wosCountThru2011)) + 
  geom_point(aes(color= journal))
p
```

It looks like most of the citation counts are close to 0. We can quickly check the distribution of this variable using a qplot.

```{r, message = FALSE}
qplot(data = research, x = wosCountThru2011)
```

To control the plot axes, we use variants of the functions `scale_x_*` and `scale_y_*`.

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = wosCountThru2011)) + 
  geom_point(aes(color= journal)) +
  scale_y_log10()
p
```

How can we fix this?

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal))
p
```

Notice what this fix has done to the way the y-axis is labelled. To manually update the axis label, we can use the `scale_y_continuous()` function.

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal)) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100, 1000), name = "Citations")
p
```

The scale family of functions can also be used to change the colours.  

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal)) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  scale_colour_manual(values = c("red","blue", "green", "yellow", "purple", "orange","cyan"), name = "Journal")
p

library(viridis)

p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal)) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  scale_colour_manual(values = viridis(7), name = "Journal")
p
```

## Colour palettes

Choosing good colours to aid visualization is not trivial and requires some (or a lot of) thought. A set of colour palettes have been developed for easy use with ggplot2 through a package called `RColorBrewer`. 

There are three types of these premade palettes:
* Sequential
* Diverging
* Qualitative

```{r, fig.height = 8}
library(RColorBrewer)
display.brewer.all()
```

Or, you can create your own color palettes using the `colorRamp()` or `colorRampPalette()` function. 

`colorRamp()` returns a function that takes values between 0 and 1, indicating the extremes of the color palette.

`colorRampPalette()` returns a function that takes integer arguments and returns a vector of colours. 

### `colorRamp()` {-}

```{r}
cols <- colorRamp(c("red", "blue"))
cols(0)
cols(0.5)
cols(1)
```

### `colorRampPalette()` {-}

```{r}
cols <- colorRampPalette(c("red", "blue"))
cols(2)
cols(10)
```

## Faceting

There are two functions to control how plots are divided into facets: `facet_wrap()` and `facet_grid()`. 

### `facet_wrap()` {-}

```{r, message = FALSE}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1),
                          color = journal)) +
  geom_point() +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  facet_wrap(~journal) +
  geom_smooth(color = "black", method = "lm")
p
```

It is still difficult to see the spread of the data. Try making the points smaller and transparent.

```{r, message = FALSE}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1),
                          color = journal)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  facet_wrap(~journal) +
  geom_smooth(color = "black", method = "lm")
p
```

### `facet_grid()` {-}

What about faceting by two variables? For example, let's say we want to look at the relationship between days since published and citations for low impact versus high impact articles.

First, let's recreate a variable categorizing high and low impact based on the variable `f1000Factor`.

```{r, message = FALSE}
research <- research %>% mutate(impact = cut(f1000Factor, breaks = c(-Inf,1,Inf), labels = c("low", "high")))

p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1),
                          color = journal)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  facet_grid(journal~impact) +
  geom_smooth(color = "black", method = "lm")
p
```

By default, R will plot each panel on the same axes. You can override this using the option `scales`, setting it to `free_x` to vary the scale across rows, `scale_y` to vary it across columns, or `free` to vary across both.

```{r, message = FALSE}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1),
                          color = journal)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  facet_wrap(~journal, scale = "free") +
  geom_smooth(color = "black", method = "lm")
p
```

---

### Exercise 5.6 {-}

Use what you've learned to generate the plot below.

```{r, echo = FALSE, message = FALSE}
ggplot(research, aes(x = log10(wosCountThru2011 + 1),
							  y = log10(pdfDownloadsCount + 1),
							  color = journal)) +
	geom_point(size = 0.5, alpha = 0.5) +
	theme_bw() +
	geom_smooth() +
	theme(legend.position = "none") +
	facet_wrap(~journal, nrow = 2) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "Citations") +
  scale_x_continuous(breaks = c(1,2,3), labels = c(10, 100,1000), name = "PDF Downloads")
```

---

## Themes

In addition to `theme_bw()`, which changes the plot background to white, `ggplot2` comes with several other themes which can be useful to quickly change the look of your visualization. The complete list of themes is available at https://ggplot2.tidyverse.org/reference/ggtheme.html. `theme_minimal()` and `theme_light()` are popular, and `theme_void()` can be useful as a starting point to create a new hand-crafted theme.

The `ggthemes` package provides a wide variety of options (including an Excel 2003 theme). The `ggplot2` [extensions website](https://www.ggplot2-exts.org) provides a list of packages that extend the capabilities of `ggplot2`, including additional themes.

This is also the place you can customize the text of your plot. 

```{r}
p <- ggplot(research, aes(x = daysSincePublished, 
                          y = log10(wosCountThru2011 + 1))) + 
  geom_point(aes(color= journal)) +
  scale_y_continuous(breaks = c(1,2,3), labels = c(10, 100, 1000), name = "Citations") +
  theme_bw() +
  xlab("Days Since Published") +
  theme(axis.title.x = element_text(size = 14, face = "bold"))
p
```

## Multiple plots

There are two useful packages for combining multiple plots: `gridExtra` and `cowplot`. 

`gridExtra` has two useful functions: `grid.arrange()` and `arrangeGrob()`. However, these functions make no attempt at aligning the plot panels; instead, the plots are simply placed into the grid as they are, so the axes are not aligned. If axis alignment is required, the `plot_grid()` function of the `cowplot` package is better. We will try using both here. 

### The `gridExtra` package {-}

Let's say we want to combine the following three plots:

```{r, message = FALSE, fig.height = 4, fig.width = 4}
plot1 <- ggplot(research, aes(x = journal, fill = journal)) +
	geom_bar(color = "black") +
	theme_bw()
plot1

plot2 <- ggplot(research, aes(x = log10(wosCountThru2011 + 1),
							  y = log10(pdfDownloadsCount + 1),
							  color = journal)) +
	geom_point(size = 0.5, alpha = 0.5) +
	theme_bw() +
	geom_smooth() +
  xlab("Citations") +
  ylab("Downloads") +
  theme(legend.position = "none")
plot2

plot3 <- ggplot(research, aes(x = impact,
							  y = log10(pdfDownloadsCount + 1),
							  fill = journal)) +
	geom_boxplot() + 
	theme_bw() +
  ylab("Downloads")
plot3
```

To arrange them together, use the `grid.arrange()` function. 

```{r, message = FALSE}
library(gridExtra)
grid.arrange(plot1, plot2, plot3)
```

To modify how they are arranged, you can use the `layout_matrix` argument.

```{r, message = FALSE}
grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2, 
             layout_matrix = rbind(c(1,2), c(3,3)))
```

### The `cowplot` package {-}

We can also arrange these plots using the `plot_grid()` function.

```{r, message = FALSE}
library(cowplot)

plot_grid(plot1, plot2, plot3)

row1 <- plot_grid(plot1, plot2, labels = c("A", "B"), nrow = 1)
row2 <- plot_grid(plot3, labels = c("C"), nrow = 1)

plot_grid(row1, row2, nrow = 2)
```

### Saving plots

The easiest way to save a plot is using the `ggsave()` function. 

---

### Exercise 8.1 {-}

Using the publication dataset we've been using, complete the following tasks. 

1. Generate a table containing the mean and standard error of the number of tweets per journal. It should look something like this:

```{r, echo= FALSE, message = FALSE}
tweets_per_journal <- research %>%
group_by(journal) %>%
summarize(num = n(),
mean = mean(backtweetsCount),
sem = sd(backtweetsCount / sqrt(num)))

tweets_per_journal
```

2. Use this table to create a bar plot of mean tweets per journal, including standard error. It should look something like this:

```{r, echo= FALSE, message = FALSE}
tweets_bar <- ggplot(tweets_per_journal, aes(x = journal, y = mean)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 0.1) +
  geom_text(aes(y = mean + sem, label = paste("N =",num)), hjust = 0.5, vjust = -0.3, size = 3) +
  theme_bw() +
  labs(y = "Mean tweets", x = "Journal")
tweets_bar
```

Hint: you will want to use the function `geom_errorbar()` to add the error bars.

<br>

3. Create a box plot of the number of citations per journal (ie. the variable wosCountThur2011). Colour the journals using the `Accent` palette from the `RColorBrewer` package. It should look something like this:

```{r, echo= FALSE, message = FALSE}
citations <- ggplot(research, aes(x = journal,
                              y = log10(wosCountThru2011 + 1),
                              fill = journal)) +
  geom_boxplot() +
  scale_y_continuous(breaks = c(1, 2,3), labels = c(10,100, 1000), name = "Citations") +
  scale_fill_brewer(palette = "Accent") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Journal")
citations
```

4. Generate a plot that shows the number of downloads over time, coloured by journal. It should look something like this:

```{r, echo= FALSE, message = FALSE}
means <- research %>%
group_by(journal, year) %>%
summarize(mean = mean(pdfDownloadsCount))

dls <- ggplot(means, aes(x = year, y = mean, colour = journal)) +
geom_line() +
# facet_wrap(~impact, nrow = 2) +
  scale_color_brewer(palette = "Accent") +  
  scale_y_continuous(breaks = c(1, 2,3), labels = c(10,100, 1000), name = "Downloads") +
  scale_x_continuous(name = "Journal") +
  theme_bw()
dls
```

---

## Additional Resources

http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization

http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html


* Mailing list: http://groups.google.com/group/ggplot2
* Wiki: https://github.com/hadley/ggplot2/wiki
* Website: http://had.co.nz/ggplot2/
* StackOverflow: http://stackoverflow.com/questions/tagged/ggplot

Cheatsheet



